---
title: "Final Project - Near-Earth Asteroids"
author: "Braxton Wilson, Richard Sauve"
date: "12/13/2021"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

* Keeping track of NEO's (near-earth objects) is important for the safety of our satellites, astronauts, and (in the case of large objects) the planet!
* Another good reason to keep track of NEO's is for the potential, in the future, to be able to mine these asteroids for minerals and water.
* Will be exploring and presenting an NEA (near-earth asteroid) dataset from Kaggle which was compiled from NASA's CNEOS website.
* Explore the data by comparing hazardous to non-hazardous NEA's and comparing different classes of NEA's.
* This analysis will help shed light on the following:
  + What are the differences between Hazardous and Non-Hazardous NEA's?
  + What are the differences between the classifications of NEA's?

## Packages Required
```{r, message=FALSE, warning=FALSE}
# several packages useful for working with dataframes (dplyr, tibbles) and graphs (ggplot2)
library(tidyverse)
# For prettier tables
library(kableExtra)
```

## Data Prep

* Data set downloaded from Kaggle:
  + https://www.kaggle.com/shrutimehta/nasa-asteroids-classification
* Data itself is from NASA
  + http://neo.jpl.nasa.gov/
* This data is public domain and was scraped from NASA's website by Shruti Mehta in 2017. She says her inspiration for collecting this data was finding potential hazardous and non-hazardous asteroids and understanding what distinguishes the two categories.
* From the NASA CNEOS website's [about](https://cneos.jpl.nasa.gov/about/) page:

  > The Center for Near-Earth Object Studies (CNEOS) computes high precision orbits for Near-Earth Objects (NEOs), predicts their future motions, assesses their impact hazard, and makes these result available on this website.
  
* No missing data for any of the variables!
* Import is simple, read.csv.
* A glimpse at the dataframe:

```{r, echo = FALSE}
# import data from csv
nea <- read.csv(file = "nasa.csv")

# select variables of interest and rename some variables
nea <- select(nea, Neo.Reference.ID, Absolute.Magnitude,
              Max.Est.Dia.KM = Est.Dia.in.KM.max.,
              Min.Est.Dia.KM = Est.Dia.in.KM.min.,
              Relative.Velocity.km.per.sec,
              Miss.Dist.AU = Miss.Dist..Astronomical.,
              Orbit.Uncertainty = Orbit.Uncertainity, Eccentricity,
              Semi.Major.Axis, Perihelion.Dist = Perihelion.Distance,
              Aphelion.Dist, Hazardous)

# change values of hazardous values as shown for plot legends
nea$Hazardous[nea$Hazardous == "True"] <- "Hazardous"
nea$Hazardous[nea$Hazardous == "False"] <- "Non-Hazardous"

# take a look at the dataframe
glimpse(nea)
```

***

### Variables of Interest:

```{r, echo = FALSE}
# Vectors for variable names and descriptions.
v_names <- c(nea %>% colnames)
descrip <- c("ID number assigned to the asteroid by NASA", "How visible the asteroid would be from a standardized distance away (1 AU) without any kind of obfuscation", "The maximum estimation of the NEA's diameter", "The minimum estimation of the NEA's diameter", "The velocity of the NEA relative to our vantage point", "Distance the object missed Earth in astronomical units", "0-9 logarithmic scale of the uncertainty of an object's orbit", "Measure of an objects deviation from a circular orbit, 0 = Perfect circle, 0 < and < 1 is an ellipse, 1 is parabolic, and > 1 is hyperbolic", "Half of the major axis of the elliptical orbit", "The point in an object's orbit closest to the sun", "The point in an object's orbit furthest from the sun", "Whether or not the asteroid meets the requirements to be deemed potentially hazardous")

# Combining these two vectors into a dataframe and renaming the columns.
var_interest <- data.frame(v_names, descrip)
colnames(var_interest) <- c("Variables", "Description")

# Displaying the variables of interest using KableExtra to set options for hover, condensed, and links to further variable information (if available, otherwise links to the CNEOS website homepage) in a table.
var_interest %>% kbl() %>% kable_minimal(c("hover", "condensed")) %>% 
  column_spec(1, color = "blue",
              link = c("https://naif.jpl.nasa.gov/pub/naif/toolkit_docs/C/req/naif_ids.html#Asteroids",
                       "https://cneos.jpl.nasa.gov/glossary/h.html", 
                       "https://cneos.jpl.nasa.gov/",
                       "https://cneos.jpl.nasa.gov/",
                       "https://cneos.jpl.nasa.gov/",
                       "https://cneos.jpl.nasa.gov/",
                    "https://minorplanetcenter.net//iau/info/UValue.html",
                  "https://cneos.jpl.nasa.gov/glossary/eccentricity.html",
                  "https://ssd.jpl.nasa.gov/glossary/semimajor_axis.html",
                  "https://cneos.jpl.nasa.gov/glossary/perihelion.html",
                  "https://cneos.jpl.nasa.gov/glossary/aphelion.html",
                  "https://cneos.jpl.nasa.gov/glossary/PHA.html"))
```

***

## Exploratory Data Analysis

* When looking through the observations available for NEA's in this data set, I wondered what made an asteroid "hazardous" or not.
* This was easily answered via NASA's CNEOS [website](https://cneos.jpl.nasa.gov/about/neo_groups.html):

  > Potentially Hazardous Asteroids (PHAs) are currently defined based on parameters that measure the asteroid's potential to make threatening close approaches to the Earth. Specifically, all asteroids with an Earth Minimum Orbit Intersection Distance (MOID) of 0.05 au or less and an absolute magnitude (H) of 22.0 or less are considered PHAs. In other words, asteroids that can't get any closer to the Earth (i.e., MOID) than 0.05 au (roughly 7,480,000 km or 4,650,000 mi) or are smaller than about 140 m (~500 ft) in diameter (i.e., H = 22.0 with assumed albedo of 14%) are not considered PHAs.

* On that same page, a graphic explaining how to calculate NEA classifications was available.

***

### NEA Classes

* Using Aphelion, Perihelion, and Absolute Magnitude a new variable Asteroid.Class was mutated.
* Calculated via the provided information from CNEOS:

![[NEA Classification](https://cneos.jpl.nasa.gov/about/neo_groups.html)](neo_orbit_types.png){width=50%}
```{r, echo=FALSE, message=FALSE}
# filtering with control statements specified by the attached image and mutating those results into separate dataframes
nea %>%
  filter(Semi.Major.Axis > 1.0, Perihelion.Dist < 1.3, Perihelion.Dist > 1.017) %>%
  mutate(Asteroid.Class = "Amors") -> Amors_asteroid
nea %>%
  filter(Semi.Major.Axis > 1.0, Perihelion.Dist < 1.017) %>%
  mutate(Asteroid.Class = "Apollos") -> Apollos_asteroid
nea %>%
  filter(Semi.Major.Axis < 1.0, Aphelion.Dist > 0.983) %>%
  mutate(Asteroid.Class = "Atens") -> Atens_asteroid
nea %>%
  filter(Semi.Major.Axis < 1.0, Aphelion.Dist < 0.983) %>%
  mutate(Asteroid.Class = "Atiras") -> Atiras_asteroid

# joining all the newly classified asteroid dataframes back into one dataframe
nea_final <- full_join(Amors_asteroid, Apollos_asteroid)
nea_final <- full_join(nea_final, Atens_asteroid)
nea_final <- full_join(nea_final, Atiras_asteroid)
```

***

### Hazardous vs Non-Hazaradous by Asteroid Class

```{r, echo = FALSE}
nea_final %>%
  select(Hazardous, Asteroid.Class) %>%
  ggplot(aes(x = Asteroid.Class, color = Asteroid.Class, fill = Asteroid.Class)) + geom_bar() + facet_wrap(~Hazardous) + labs(x = "Asteroid Class", y = "Count", color = "Asteroid Class", fill = "Asteroid Class")
```


```{r, echo=FALSE}
# table of percentages of Hazardous NEA's from each class
ast <- c("Amors", "Apollos", "Atens", "Atiras")
per <- c("2.79%", "73.52%", "23.02%", "0.67%")
df <- data.frame(ast, per)
colnames(df) <- c("Asteroid Class", "Percentage of Hazardous NEA's from each class")
df %>% kbl() %>% kable_minimal("striped")
```

```{r, echo=FALSE}
# table of percentages of each NEA class that is hazardous
per <- c("2.23%", "22.13%", "15.10%", "35.71%")
df <- data.frame(ast, per)
colnames(df) <- c("Asteroid Class", "Percentage of NEA Class that is Hazardous")
df %>% kbl() %>% kable_minimal("striped")
```

***

### Absolute Magnitude

* Based off the previous definition of a potentially hazardous asteroid, the average absolute magnitude of hazardous asteroids should be lower than that of non-hazardous asteroids.
* The five largest absolute magnitudes:
```{r, echo = FALSE}
# Arranging and selecting for the 5 largest Absolute Magnitudes piped into a KableExtra table
nea_final %>% arrange(desc(Absolute.Magnitude)) %>% select(Neo.Reference.ID, Absolute.Magnitude, Hazardous) %>% head(5) %>% kbl() %>% kable_minimal(full_width = F, c("striped", "hover", "condensed"))
```
* The five smallest absolute magnitudes:
```{r, echo = FALSE}
# Arranging and selecting for the 5 smallest Absolute Magnitudes piped into a KableExtra table (cannot for the life of me figure out why tail kept the indices while, above, head did not)
nea_final %>% arrange(desc(Absolute.Magnitude)) %>% select(Neo.Reference.ID, Absolute.Magnitude, Hazardous) %>% tail(5) %>% kbl() %>% kable_minimal(full_width = F, c("striped", "hover", "condensed"))
```

***

```{r, echo = FALSE}
# A 4 plot of densities separated by asteroid class for hazardous and non-hazardous with a vertical dotted line drawn at Absolute Magnitude (x) = 22
nea_final %>%
  ggplot(aes(x = Absolute.Magnitude, color = Hazardous, fill = Hazardous)) + geom_density(alpha = 0.5) + facet_wrap(~Asteroid.Class) + geom_vline(xintercept = 22, size = .5, color="black", linetype=2) + annotate(geom = "text", label = "H = 22", x = 22.5, y=0.22, angle = 90, size=2.5) + labs(x = "Absolute Magnitude (H)", y = "Density", title = "NEA Absolute Magnitude")
```

***

* Summary Statistics

```{r, echo = FALSE}
# Print absolute magnitude summary statistics separated by hazard classification.
print("Hazardous:")
summary(nea_final$Absolute.Magnitude[nea_final$Hazardous=="Hazardous"])
print("Non-Hazardous:")
summary(nea_final$Absolute.Magnitude[nea_final$Hazardous=="Non-Hazardous"])
```

* As expected and shown in the density graphs, the absolute magnitude of non-hazardous NEA's is larger than that of hazardous NEA's and non-hazardous NEA's have a much wider range.

***

### Orbital Uncertainty

* A look at the differences between Hazardous/Non-Hazardous NEA's and NEA classes of Orbital Uncertainty.

```{r, echo = FALSE, warning = FALSE}
# Box plot of Orbital Uncertainty by Hazard and split by Asteroid Class to show how much more uncertain the orbits of non-hazardous NEA's are compared to hazardous NEA's.
nea_final %>%
  ggplot(aes(x = Orbit.Uncertainty, y = Hazardous, fill = Hazardous)) + geom_boxplot() + facet_wrap(~Asteroid.Class) + labs(title = "Orbital Uncertainty by Hazard and Asteroid Class", x = "Orbital Uncertainty")
```

* Hazardous NEA's have significantly less orbital uncertainty on average than non-hazardous NEA's excluding Atiras class NEA's.
* Part of the Orbital Uncertainty parameter's calculation is the number of observations, so it very well may be the case that astronomers spend more time observing potentially hazardous asteroids.
* There is very little orbital uncertainty in Atiras class NEA's.

***

### NEA Size & Velocity

* From the definition of PHA's, PHA's should be larger on average than non-hazardous NEA's.

```{r, echo = FALSE}
nea_final %>% ggplot(aes(x = Relative.Velocity.km.per.sec,
                         y = Max.Est.Dia.KM, color = Hazardous)) + geom_density_2d() + facet_wrap(~Asteroid.Class) + labs(title = "Relative Velocity (km/s) vs Maximum Estimated Diameter (km)", x = "Relative Velocity (km/s)", y = "Maximum Estimated Diameter (km)")
```


* Our expectation is correct, hazardous NEA's are larger on average than non-hazardous NEA's. Again, much larger range for non-hazardous asteroids on both ends of the spectrum.
* The mean, being more subject to outliers, is significantly different than the median for both hazard classifications of asteroid sizes.
* As expected, hazardous NEA's have a much greater relative velocity on average than non-hazardous NEA's.

***

### The Largest Asteroid

* After seeing the tremendous right skew in the previous boxplot, let's take a look at the largest NEA's:

```{r, echo = FALSE}
# table to view the 5 largest NEA's in this dataset
nea_final %>% select(Neo.Reference.ID, Max.Est.Dia.KM, Min.Est.Dia.KM, Hazardous) %>% arrange(desc(Max.Est.Dia.KM)) %>% head(5) %>% kbl() %>% kable_minimal("hover")
```

* 2000433 is the largest by far and is the culprit of the horrible skew in the previous boxplot.

#### 433 Eros
![[433 Eros](https://solarsystem.nasa.gov/asteroids-comets-and-meteors/asteroids/433-eros/in-depth/)](454_Eros_main.jpg){width=50%}

* 433 Eros is the first asteroid to be orbited by a satellite and the first to have a satellite land on its surface (NEAR Shoemaker).
* This is an actual photo from NASA's CNEOS website taken by NEAR Shoemaker.
* 433 Eros is of NEA class Amors and ended up being slightly smaller than expected: 34 km x 11.2 km x 11.2 km 
* The asteroid (or comet) that wiped out the dinosaurs is estimated to be 10 km in diameter...
* Thank goodness 433 Eros isn't Hazardous!

***

### Perihelion Distance and Eccentricity

```{r, echo = FALSE}
# Scatterplot of perihelion distance vs eccentricity colored by hazard classification
nea_final %>% ggplot(aes(x = Perihelion.Dist, y = Eccentricity, color = Hazardous, fill = Hazardous)) + geom_point() + labs(title = "Perihelion Distance vs Eccentricity", x = "Perihelion Distance (au)")

# Scatterplots of perihelion distance vs eccentricity colored by hazard classification and separated by asteroid class
nea_final %>% ggplot(aes(x = Perihelion.Dist, y = Eccentricity, color = Hazardous, fill = Hazardous)) + geom_point() + facet_wrap(~Asteroid.Class)+ labs(title = "Perihelion Distance vs Eccentricity by Asteroid Class", x = "Perihelion Distance (au)")
```

* A nice visual display of how perihelion distance is used to calculate asteroid class.
* Also a nice visual of the relationship between eccentricity and perihelion distance.
* The greater the perihelion distance the lower the eccentricity which is due to eccentricity being used to calculate perihelion distance:
  + P = semi-major axis * (1 - eccentricity).
* This graph also displays the perihelion cut-off for PHA's (a little above 1 AU).

***

### Miss Distance

* How close do these objects get to the Earth?

```{r, echo = FALSE}
# Graph Miss Distance by Hazard Class separated by Asteroid Class
nea_final %>% ggplot(aes(x = Miss.Dist.AU, y = Hazardous, fill = Hazardous)) + geom_boxplot() + facet_wrap(~Asteroid.Class) + labs(title = "Miss Distance in Astronomical Units", x = "Miss Distance (au)")

# Summary Statistics on Miss Distances by hazard class
print("Hazardous NEA Miss Distance (au) Summary Statistics")
summary(nea_final$Miss.Dist.AU[nea_final$Hazardous=="Hazardous"])
print("Non-Hazardous NEA Miss Distance (au) Summary Statistics")
summary(nea_final$Miss.Dist.AU[nea_final$Hazardous!="Hazardous"])
```

* On average non-hazardous NEA's miss by less than hazardous NEA's.
  + Looks like a small difference but it's important to remember that these are measured in astronomical units (1 au is approximately 149,598,000 km).
* Non-hazardous NEA's are smaller on average so many of them wouldn't be classified as potentially hazardous even though they miss earth by less distance on average.

***

## Summary

* Problem Statements:
  + What are the differences between Hazardous and Non-Hazardous asteroids?
  + What are the differences between the classifications of NEA’s?
* These questions were answered by comparing the selected variables separated by hazard and NEA classification via graphs and basic statistics.
* Interesting Insights
  + NEA's are classified by orbit (inside earth's orbit, intersecting earth's orbit but mostly inside, intersecting earth's orbit but mostly outside, and outside earth's orbit)
  + Most NEA's are Amors, Apollos, Atens class.
    - Most hazardous asteroids are Apollo or Atens class (over 96%).
    - There are very few Atiras class NEA's.
  + Orbital uncertainty is very low in Hazardous NEA's (mean is 0!)
    - As mentioned earler, this is probably because we spend more time observing PHA's
  + Hazardous asteroids are less visible and larger on average (as expected from how PHA's are defined).
    - The largest NEA in this data set (433 Eros) is non-hazardous.
  + Hazardous asteroids have a significantly greater relative velocity.
* Implications of our findings
  + We are more aware of the celestial bodies in our neighborhood than ever before.
  + This should allow us to be better prepared to avoid potentially devastating impacts and to keep our astronauts & satellites safe.
  + As our technology advances, humans should have a better idea of which asteroids to mine (once/if that becomes possible).
* Limitations
  + Would have liked if there had been an estimated mass variable for these NEO's. This would have opened up more possibilities for more calculations (gravity, momentum, etc.).
  + NASA's CNEOS has found significantly more NEO's since this dataset was compiled (433 Eros is no longer the [largest NEA](https://neo.ssa.esa.int/search-for-asteroids?sum=1&des=1036%20Ganymed#_NEOSearch_WAR_PSDBportlet_:j_idt10:j_idt583:physicalPropertiesTab)).
  + A big limitation is my lack of knowledge of physics and astronomy, being able to work with an astronomer would have been a big help.
  + More information on asteroid composition would be nice but advances in technology are necessary for this to be available.
